---
title: "Keep an R workflow on the GPU"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Keep an R workflow on the GPU}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A GPU is most useful when data are not repeatedly copied between R and the
GPU. This guide shows the simple pattern: upload once, run several CUDA tasks,
and bring back only the result you need.

## Require CUDA at the start

Use a strict selection before a long analysis:

```{r select}
library(cudaverse)
cuda_select_device("cuda")
```

If CUDA is not ready, this call stops and explains what is missing. It never
turns a requested GPU workflow into an unnoticed non-GPU run.

## Upload once and reuse the tensor

Construct a `cudatensor` once, then apply arithmetic, reshape, transpose,
subsetting, reduction, and matrix multiplication without reconstructing the
input for every call.

```{r tensor-pipeline}
set.seed(1)
x <- matrix(rnorm(10000 * 100), nrow = 10000)
x_gpu <- cuda_tensor(x, device = "cuda", dtype = "float32")

means_gpu <- tensor_mean(x_gpu, dim = 1)
centered_gpu <- x_gpu - means_gpu
selected_gpu <- centered_gpu[, 1:50, drop = FALSE]
gram_gpu <- tensor_matmul(t(selected_gpu), selected_gpu)

tensor_device(gram_gpu)
cuda_provenance(gram_gpu)
```

Supported reshape and selection operations stay on the GPU, so they can feed
the next calculation without downloading the full tensor.

Only cross the host boundary when an ordinary R object is required:

```{r tensor-download}
gram <- to_cpu(gram_gpu)
```

Calling `to_cpu()` inside each stage would add transfers and erase much of the
reason to use the GPU.

## Keep PCA followed by kNN on CUDA

The native PCA object retains device storage for its scores, so exact kNN can
consume `pca$x` directly.

```{r pca-knn}
pca <- cuda_pca(
  x_gpu,
  n_components = 20,
  device = "cuda"
)

neighbors <- cuda_knn(
  pca$x,
  k = 15,
  batch_size = 256,
  device = "cuda"
)

cuda_provenance(pca)
cuda_provenance(neighbors)
head(neighbors$index)
```

The public kNN result contains ordinary R index and distance matrices because
those are normally the final analysis output. The expensive PCA, distance
blocks, and stable top-k stages remain CUDA stages.

## Keep sparse preprocessing on CUDA

For sparse data, upload a `Matrix` object once and continue with normalization,
PCA, and exact kNN:

```{r sparse-resident}
counts <- Matrix::rsparsematrix(50000, 128, density = 0.01)
counts@x <- abs(counts@x)

counts_gpu <- cuda_sparse(counts, device = "cuda")
normalized_gpu <- sparse_normalize(
  counts_gpu,
  margin = "rows",
  scale_factor = 10000,
  log1p = TRUE
)
sparse_pca <- cuda_pca(normalized_gpu, n_components = 20, device = "cuda")
sparse_neighbors <- cuda_knn(
  sparse_pca$x,
  k = 15,
  device = "cuda"
)

sparse_info(normalized_gpu)
cuda_provenance(sparse_neighbors)
```

Use `to_dgCMatrix()` only when a downstream R package specifically needs a
host `dgCMatrix`.

## Read provenance

`cuda_provenance()` shows where each part of a calculation ran. The most useful
fields are:

| Field | Question answered |
|---|---|
| `operation` | What did this stage do? |
| `device` | Where did the computation run? |
| `backend` | Which cudaverse implementation performed it? |
| `output_device` | Where was the result left for the next step? |

Multi-stage algorithms retain a stage list. Inspect it directly when checking
residency:

```{r inspect-stages}
prov <- cuda_provenance(neighbors)
prov$stages
```

A strict `device = "cuda"` request cannot silently fall back. A hybrid
high-level workflow can still contain a deliberately documented host stage;
that stage has its own record rather than being described as GPU-resident.

## Control memory and transfer costs

```{r memory}
cuda_memory_info("cuda")
```

- Prefer `float32` when its numerical precision is sufficient; it halves the
  value storage compared with `float64`.
- Reuse `cudatensor` and `cudasparse` objects across operations.
- Use `cuda_knn()` instead of `cuda_distance()` when a full dense distance
  matrix is not needed.
- Lower `batch_size` when exact distance or kNN blocks approach the available
  device memory.
- Print or download large CUDA objects only when the host values are needed.

## Confirm the backend used

The lightweight path does not require R `torch`. A typical native result
should show CUDA and native in its stage record:

```{r assert-native}
stages <- cuda_provenance(neighbors)$stages
vapply(stages, `[[`, character(1), "device")
vapply(stages, `[[`, character(1), "backend")
```

Some records describe preparation or returning the final R object rather than
a numerical calculation. The [operation coverage guide](backend-support.html)
explains what to expect for each task.
