---
title: "Backend support and release boundaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backend support and release boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`cudaverse` has one public API and three possible computation backends. The
portable base backend is always present. The optional torch backend uses a
CUDA-enabled torch installation. The lightweight native backend dynamically
loads an NVIDIA driver, cuBLAS, cuSOLVER, and package PTX only when CUDA is
diagnosed or requested.

Installing the package does not download a CUDA runtime or LibTorch. Explicit
`device = "cuda"` requests are strict and never silently fall back to CPU.
`device = "auto"` may fall back, and the reason is preserved in provenance.

## Read the matrix

The table below covers every exported function. Its status vocabulary is:

- `direct`: the main numerical operation is implemented by that backend;
- `hybrid`: meaningful stages are split between that backend and CPU code;
- `cpu_only`: the operation deliberately runs on CPU for the 0.2 surface;
- `metadata`: validation or metadata without a new numerical kernel;
- `probe`: backend diagnosis or selection; and
- `host`: an explicit host materialization.

`direct` does not imply that the public return value remains on the device.
The `result_location` column states whether the result is host-resident,
device-resident, cached for a native continuation, or backend-dependent.

```{r capability-matrix, echo = FALSE}
matrix_path <- system.file(
  "reports", "backend-capability-matrix.csv", package = "cudaverse"
)
if (!nzchar(matrix_path)) {
  matrix_path <- file.path(
    dirname(knitr::current_input(dir = TRUE)),
    "..", "inst", "reports", "backend-capability-matrix.csv"
  )
}
support <- utils::read.csv(matrix_path, stringsAsFactors = FALSE)
knitr::kable(support, row.names = FALSE)
```

The installed matrix and `NAMESPACE` are checked together: a new or removed
export fails the package tests until this table is updated.

## Intentional 0.2 CPU boundaries

The following boundaries are deliberate and are not reasons to install more
packages:

- native CUDA tensor subsetting and replacement keep tensor values on the
  device; missing indices and compatibility backends without indexing
  operations use a recorded CPU round trip;
- `t()` preserves `cudasparse` device, format, and labels; native backing
  storage is transposed on CUDA while compatibility backends rebuild from the
  stable public COO metadata;
- native CUDA k-means keeps distance, stable assignment, accumulation, and
  centre updates on the device; the torch compatibility path uses GPU distance
  and CPU assignment/centre updates;
- kNN graph construction uses `Matrix`, while Louvain and Leiden use `igraph`;
- UMAP and t-SNE use `uwot` and `Rtsne` on CPU; and
- diffusion maps may calculate distance on CUDA, but construct the kernel and
  eigendecomposition on CPU.

Use `cuda_provenance()` on a result to inspect each actual compute and transfer
stage. A hybrid result is not presented as fully GPU-resident.

## Native acceptance gate

On 2026-08-10, source commit
`fdceea66d13c84c4adf731b8cd4fd0826c738ca6` passed the complete explicitly
gated native test file on an NVIDIA RTX 2000 Ada Generation GPU with 16 GB VRAM
and driver 595.97. No hardware case was skipped. The gate covered:

- float32/float64 transfer, matmul, arithmetic, reductions, reshape,
  broadcasting, transpose, subsetting, and replacement;
- SVD, PCA, PCA prediction, distance, deterministic top-k, and kNN;
- COO/CSR ownership, sparse multiplication, reductions, normalization, sparse
  PCA, and sparse kNN;
- structured CUDA and R error recovery, time-limit interruption, and backend
  reuse; and
- shared ownership, allocation telemetry, and 1,000-cycle dense and sparse
  lifecycle tests.

The test process referenced user-provided `cublas64_12.dll` and
`cusolver64_11.dll` files already present on that machine. These libraries are
not copied into or redistributed with `cudaverse`. Exact release-candidate
evidence is regenerated at the final candidate commit rather than inferred
from this checkpoint.

## Inspect a local installation

```{r diagnostics}
library(cudaverse)

diagnostics <- cuda_diagnostics()
diagnostics$selected_backend
diagnostics$backend_diagnostics$native$capabilities
diagnostics$backend_diagnostics$native$operations
```

`capabilities` are user-facing feature declarations. `operations` are the
callable internal registry entries. Native automatic selection additionally
requires the compatible contract schema, complete runtime components, and a
passing cached self-test.
