---
title: "What runs on CUDA?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What runs on CUDA?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Use this page to decide whether cudaverse can accelerate the part of your
analysis that matters. “Yes” means the numerical work runs with CUDA. “Partly”
means cudaverse uses CUDA for some steps and an established R package for the
rest.

## CUDA task summary

| Task | Runs with CUDA? | What you receive in R |
|---|---|---|
| Dense tensor operations | Yes | A GPU tensor until you call `to_cpu()` |
| Matrix multiplication and summaries | Yes | A GPU tensor |
| SVD and PCA | Yes | An R model that can reuse GPU results |
| Pairwise distances | Yes | A regular R distance matrix |
| Exact kNN | Yes | Neighbour indices and distances |
| k-means | Yes | An R clustering result |
| Sparse storage, normalization, and multiplication | Yes | A GPU sparse object |
| Sparse PCA followed by exact kNN | Yes | Final neighbour results |
| kNN graph construction | Partly | A sparse `Matrix` graph |
| Louvain and Leiden | Partly | Community labels |
| UMAP and t-SNE | Partly | Embedding coordinates |
| Diffusion map | Partly | Embedding coordinates |

Use `device = "cuda"` for tasks that must run on an eligible CUDA backend and
inspect `cuda_provenance()` for the actual stage boundary.

## Dense CUDA example

```{r dense}
library(cudaverse)
cuda_select_device("cuda")

set.seed(1)
x <- matrix(rnorm(10000 * 100), nrow = 10000)
x_gpu <- cuda_tensor(x, device = "cuda", dtype = "float32")

product_gpu <- tensor_matmul(t(x_gpu), x_gpu)
reduced_gpu <- tensor_sum(product_gpu, dim = 1)
pca <- cuda_pca(x_gpu, n_components = 20, device = "cuda")
neighbors <- cuda_knn(pca$x, k = 15, device = "cuda")

tensor_device(reduced_gpu)
cuda_provenance(neighbors)
```

The tensor operations and PCA scores stay on the GPU. Only the final neighbour
matrices are returned as ordinary R objects.

## Sparse CUDA example

```{r sparse}
counts <- Matrix::rsparsematrix(10000, 100, density = 0.03)
counts@x <- abs(counts@x)

counts_gpu <- cuda_sparse(counts, device = "cuda")
normalized_gpu <- sparse_normalize(
  counts_gpu,
  margin = "rows",
  scale_factor = 10000,
  log1p = TRUE
)
feature_totals_gpu <- sparse_row_sums(t(normalized_gpu))
sparse_pca <- cuda_pca(normalized_gpu, n_components = 20, device = "cuda")
sparse_neighbors <- cuda_knn(
  sparse_pca$x,
  k = 15,
  device = "cuda"
)

sparse_info(normalized_gpu)
cuda_provenance(sparse_neighbors)
```

## Graph and embedding workflows

GPU acceleration of PCA and exact kNN can still reduce the cost before graph
or embedding stages, but the current graph clustering and UMAP/t-SNE
implementations are not described as native CUDA.

```{r graph-embedding}
pca <- cuda_pca(x, n_components = 20, device = "cuda")
neighbors <- cuda_knn(pca$x, k = 15, device = "cuda")

graph <- cuda_knn_graph(neighbors)
communities <- cuda_leiden(graph)
embedding <- cuda_umap(pca$x)

cuda_provenance(communities)
cuda_provenance(embedding)
```

The provenance record shows the CUDA preprocessing stages and the intentional
host stages separately. This makes it possible to use the accelerated portion
without presenting the complete workflow as GPU-resident.

## The lightweight backend

The native backend loads the NVIDIA libraries already installed on your
computer and does not require LibTorch. The selected backend is always visible
in diagnostics and provenance.

```{r backend-check}
diagnostics <- cuda_diagnostics()
diagnostics$selected_backend
diagnostics$backend_status
```

An explicit CUDA request is strict: if CUDA is not ready, cudaverse stops and
shows diagnostic guidance instead of silently changing where the work runs.
