---
title: "Backend selection and compute provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backend selection and compute provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`cudaverse` separates four questions that are easy to conflate:

1. What device did the caller request?
2. Where did each stage actually compute?
3. Which concrete backend implemented that stage?
4. On which device does the returned value live?

The answers are recorded with the `cudaverse-stage/1` provenance schema.
This tutorial runs completely on CPU. Its CUDA section is optional and runs
only when a registered native or torch CUDA backend is usable.

## Inspect the runtime

```{r}
library(cudaverse)

diagnostics <- cuda_diagnostics()
diagnostics
```

`cuda_diagnostics()` never installs torch or a CUDA runtime.
`available_backends`, `auto_eligible_backends`, `auto_selection_reason`,
`selected_backend`, and `backend_diagnostics` describe the registry. The first
native diagnostic in a session runs and caches a small transfer, float32 and
float64 matmul/reduction, arithmetic/broadcast/reshape/transpose, and sparse
normalization self-test. The legacy fields
`torch_installed`, `torch_version`, `cuda_available`, `cuda_device_count`,
`reason`, and `detection_error` remain available throughout the 0.3
compatibility cycle.

Native is eligible for `"auto"` only when all four gates pass:

1. backend contract schema `cudaverse-backend/1` matches;
2. the complete tensor and resident PCA/kNN capability set is present;
3. driver, cuBLAS, cuSOLVER, and package PTX components load; and
4. the cached runtime self-test passes without retaining tracked device memory.

If any gate fails, diagnostics preserve its reason and automatic selection
tries the torch compatibility backend before using the observable CPU fallback.
Setting `options(cudaverse.cuda_backends = ...)` can constrain test order, but
cannot bypass these gates.

Device requests have these semantics:

| Requested device | Actual device | Output device at construction | Fallback |
|---|---|---|---|
| `"cpu"` | CPU | CPU | `FALSE` |
| `"auto"` with usable CUDA | CUDA | CUDA | `FALSE` |
| `"auto"` without usable CUDA | CPU | CPU | `TRUE`, with a recorded reason |
| `"cuda"` | CUDA, or an error | CUDA when successful | Never silent |

An explicit CUDA request is strict. If CUDA is unavailable,
`cuda_tensor(..., device = "cuda")` signals a
`cudaverse_cuda_unavailable` condition instead of returning a CPU tensor.

## A portable CPU workflow

Use `device = "cpu"` when reproducible portable execution matters more than
automatic hardware selection.

```{r}
left_source <- matrix(
  1:6,
  nrow = 2,
  dimnames = list(
    sample = c("sample_a", "sample_b"),
    feature = c("gene_a", "gene_b", "gene_c")
  )
)
right_source <- matrix(
  seq_len(6) / 10,
  nrow = 3,
  dimnames = list(
    feature = c("gene_a", "gene_b", "gene_c"),
    component = c("component_1", "component_2")
  )
)

left <- cuda_tensor(left_source, device = "cpu", dtype = "float64")
right <- cuda_tensor(right_source, device = "cpu", dtype = "float64")
product <- tensor_matmul(left, right)

tensor_device(product)
to_cpu(product)
```

Every returned `cudatensor` records the actual stage:

```{r}
construction_provenance <- cuda_provenance(left)
construction_provenance

product_provenance <- cuda_provenance(product)
product_provenance

attr(product_provenance, "schema")
attr(product_provenance, "compute_device")
```

The provenance table has one row per stage:

- `requested_device` is the caller's policy (`"cpu"`, `"cuda"`, `"auto"`,
  `"fixed-cpu"`, or `"inherited"`);
- `device` is where that stage actually computed;
- `backend` names the implementation, such as `base`, `native`, or `torch`;
- `selection_reason` explains the resolution;
- `fallback` is only true when an automatic request selected CPU;
- `output_device` is where that stage placed its result.

Do not infer the output location from `device`. A transfer or hybrid operation
can compute on one device and return data on another.

## Automatic selection is observable

The following call is portable. It chooses CUDA only when diagnostics report a
usable device; otherwise it returns a CPU tensor and records the fallback.

```{r}
automatic <- cuda_tensor(matrix(1:4, 2), device = "auto")
cuda_provenance(automatic)
```

This makes an automatic CPU result distinguishable from an explicit CPU
request. For production runs that require a GPU, use `device = "cuda"` rather
than `"auto"`.

## Optional CUDA path

```{r}
if (cuda_available()) {
  gpu <- cuda_tensor(left_source, device = "cuda", dtype = "float64")
  gpu_result <- tensor_sum(gpu, dim = 1)

  tensor_device(gpu_result)
  cuda_provenance(gpu_result)

  # Native CUDA gathers values on-device. A compatibility backend may record
  # materialization and upload stages instead.
  gpu_subset <- gpu[1, , drop = FALSE]
  cuda_provenance(gpu_subset)
}
```

The guard keeps this installed vignette runnable on machines without NVIDIA
hardware. It is not evidence that the CUDA path was tested.

## Dense memory and transfer limits

Dense `cudaverse` tensors use dense storage. The payload alone is approximately
`prod(shape) * bytes_per_value`, before outputs and temporary tensors. The CPU
fallback uses ordinary R vectors, so its physical storage can differ from CUDA
backends:

| dtype | CPU base storage | CUDA native storage | CUDA torch storage |
|---|---:|---:|---:|
| `float32` | 8 bytes | 4 bytes | 4 bytes |
| `float64` | 8 bytes | 8 bytes | 8 bytes |
| `integer` | 4 bytes | 4 bytes | 8 bytes |

CPU `float32` values are rounded to IEEE single precision when stored and
after each public operation, but base R kernels accumulate in double
precision. CUDA kernels compute in their backend's native dtype. Results should
therefore be compared with an explicit tolerance rather than expected to be
bit-for-bit identical. Floating tensors may contain IEEE `Inf`, `-Inf`, and
`NaN`; integer tensors reject values that cannot be represented exactly.

Creating a CUDA tensor from an R object transfers the dense input from host
memory to GPU memory. `to_cpu()`, `as.array()`, and `as.matrix()` materialize
the complete result in host memory. Mixed-device binary operations move the
right operand to the left operand's device. Native CUDA subsetting and
replacement evaluate index metadata in R but gather or replace tensor values
on the GPU. Missing indices and backends without indexing operations use a
recorded CPU round trip. Matrix arithmetic, broadcasting, reductions, reshape,
and transpose use the tensor's backend.

Large operations can temporarily require input, output, and intermediate
allocations at once. The package does not currently provide out-of-core
execution or automatic chunking. Printing a tensor with more than 100 values
shows metadata without copying the full CUDA allocation; explicitly call
`to_cpu()` only when that transfer is intended.

The built-in native pipeline stores COO and CSR metadata in shared device
storage. Sparse normalization remains sparse; sparse-input PCA expands directly
into a GPU dense buffer, attaches shared device storage to PCA's ordinary R
score matrix, and a following `cuda_knn()` call reuses that allocation. Exact
distance blocks and stable top-k selection remain on the GPU, and only compact
results are returned. `cuda_provenance()` records normalization,
`sparse_to_dense`, `scores_resident`, device-resident input materialization,
CUDA distance, and final CPU output as separate stages.

## Hardware-gated validation

Portable examples and ordinary R CMD check jobs use CPU and may guard optional
CUDA code with `if (cuda_available())`. The repository's `cuda-parity`
workflow is different: manual dispatch, or the organization variable
`CUDAVERSE_NVIDIA_CI=enabled`, invokes a reusable job on a self-hosted runner
labeled `cuda`.

The main-package compatibility job:

- sets `CUDAVERSE_REQUIRE_CUDA=true`;
- requires `nvidia-smi` to succeed;
- requires torch to report at least one CUDA device;
- runs required CPU/CUDA parity and provenance checks.

The integrated native workflow compiles `cudaverse` on Windows, macOS, Ubuntu,
and R-devel; reproducibly rebuilds its PTX in CUDA 12.8.1; and runs explicit
native parity, structured-error, interruption, and 1,000-cycle VRAM contracts
on the RTX 2000 development machine. The ordinary package checks do not need
CUDA hardware or a CUDA toolkit.

A missing CUDA runtime fails the hardware job. It is never converted into a
successful skip.
