---
title: "Getting started with cudaverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cudaverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`cudaverse` is a single entry point for dense tensors, sparse matrices,
numerical algorithms, graphs, and embeddings. CUDA is optional: every result
records the backend and device that actually performed each stage.

## Dense and sparse data

```{r}
library(cudaverse)

dense <- cuda_tensor(matrix(1:12, nrow = 4), device = "cpu")
tensor_shape(dense)

sparse <- cuda_sparse(Matrix::Diagonal(4), device = "cpu")
sparse_info(sparse)
```

Sparse normalization keeps the zero pattern and composes with the same PCA and
kNN entry points:

```{r}
counts <- matrix(c(1, 0, 3, 2, 4, 1, 0, 2, 5, 1, 3, 2), nrow = 4)
sparse_counts <- cuda_sparse(counts, device = "cpu")
normalized <- sparse_normalize(
  sparse_counts, margin = "rows", scale_factor = 100, log1p = TRUE
)
sparse_pca <- cuda_pca(normalized, n_components = 2, device = "cpu")
sparse_neighbors <- cuda_knn(sparse_pca$x, k = 2, device = "cpu")
```

## Algorithms compose directly

```{r}
set.seed(1)
x <- matrix(rnorm(120), nrow = 30)

pca <- cuda_pca(x, n_components = 4, device = "cpu")
neighbors <- cuda_knn(pca$x, k = 4, device = "cpu")
graph <- cuda_knn_graph(neighbors)

dim(pca$x)
dim(as_adjacency_matrix(graph))
cuda_provenance(pca)
```

UMAP, t-SNE, Louvain, and Leiden use optional packages. Their functions report
a clear installation message when the relevant optional backend is missing.

## Optional single-cell inputs

When `SingleCellExperiment` is installed, `cuda_umap()`, `cuda_tsne()`, and
`cuda_diffusion_map()` can consume one of its reduced dimensions directly.
Select the dimension explicitly unless there is exactly one reduced dimension
named `"PCA"`:

```{r, eval=FALSE}
fit <- cuda_diffusion_map(
  sce,
  reduced_dim = "PCA",
  n_components = 2,
  device = "auto"
)
```

The selected reduced dimension is materialized as a host matrix. UMAP and
t-SNE then run on their documented CPU backends. Diffusion maps may upload the
matrix for the distance stage, while kernel construction and eigendecomposition
remain on CPU; `cuda_provenance(fit)` reports this boundary. Seurat objects are
not currently a public input type, so pass a finite matrix or a
`SingleCellExperiment` reduced dimension instead. These optional integrations
do not add Bioconductor or Seurat to an ordinary cudaverse installation.
