---
title: "GPU setup and troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPU setup and troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`cudaverse` always installs with a portable base R backend. CUDA is optional,
runtime-lazy, and never downloaded by the package. The preferred CUDA path is
the lightweight native backend; R torch remains an optional compatibility
backend.

## Start with the CPU contract

Install the current development line and prove the same public API works on
CPU before changing the runtime:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("cudaverse/cudaverse@develop/0.4")
```

```{r cpu-contract}
library(cudaverse)

x <- matrix(seq_len(20), nrow = 5)
fit <- cuda_pca(x, n_components = 2, device = "cpu")
tensor <- cuda_tensor(x, device = "cpu")

fit$device
tensor_device(tensor)
cuda_provenance(fit)
```

If this fails, diagnose the package or input before diagnosing CUDA. Include
the CPU result in a GPU bug report.

## Native CUDA runtime

The package already contains its own checksum-pinned PTX kernels. It does not
bundle or download NVIDIA runtime libraries. Native execution discovers these
compatible system components only when CUDA is diagnosed or requested:

- an NVIDIA driver and CUDA Driver API;
- cuBLAS 12; and
- cuSOLVER 11.

A complete CUDA Toolkit is not required to install, build, or use the CPU
backend. Installing an NVIDIA distribution that provides the required runtime
libraries is one way to make native CUDA available. Follow NVIDIA's official
archived guides for the currently tested CUDA 12.8 family on
[Windows](https://docs.nvidia.com/cuda/archive/12.8.0/cuda-installation-guide-microsoft-windows/)
or [Linux](https://docs.nvidia.com/cuda/archive/12.8.0/cuda-installation-guide-linux/)
instead of copying an unverified command from an old article. Compatibility
is determined by diagnostics and the runtime self-test, not by a directory
name alone.

On Windows or Linux, explicitly point cudaverse at compatible libraries when
they are not already visible to the operating-system loader. Set these before
loading the package:

```{r native-paths, eval = FALSE}
Sys.setenv(
  CUDAVERSE_CUBLAS_PATH = "C:/path/to/cublas64_12.dll",
  CUDAVERSE_CUSOLVER_PATH = "C:/path/to/cusolver64_11.dll"
)
```

On Linux, use absolute paths to `libcublas.so.12` and `libcusolver.so.11`
instead. These variables identify user-provided runtime files; cudaverse does
not copy or redistribute them. Native CUDA is unsupported on macOS, where the
portable CPU backend remains available.

## Read diagnostics before selecting a device

Run diagnostics in a fresh R session after changing a driver or library path:

```{r diagnostics}
diagnostics <- cuda_diagnostics()
diagnostics$status
diagnostics$summary
diagnostics$next_steps
diagnostics$backend_status
diagnostics$available_backends
diagnostics$auto_eligible_backends
diagnostics$selected_backend
diagnostics$auto_selection_reason
```

For the native backend, inspect the complete gate rather than only device
count:

```{r native-diagnostics}
native <- diagnostics$backend_diagnostics$native
native[c(
  "available", "reason", "driver_version", "cublas_loaded",
  "cusolver_loaded", "kernels_loaded", "runtime_complete",
  "capability_compatible", "self_test_passed", "auto_eligible"
)]
native$self_test
native$missing_auto_capabilities
```

Native is eligible for automatic selection only when the versioned
`cudaverse-backend/1` contract, required operations, runtime components, and
cached self-test all pass. A detected GPU alone is insufficient.
When CUDA is unavailable, `next_steps` maps the stable selection reason to a
concrete action. The strict `cudaverse_cuda_unavailable` condition retains the
same reason, diagnostics, and guidance for programmatic error handling.

## Prove strict native execution

Use an explicit CUDA request when the workflow must run on a GPU. To test
native without allowing the torch compatibility backend, constrain the
session's backend order temporarily:

```{r strict-native, eval = FALSE}
old <- options(cudaverse.cuda_backends = "native")

x_gpu <- cuda_tensor(
  matrix(rnorm(1024^2), 1024),
  device = "cuda",
  dtype = "float32"
)
y_gpu <- tensor_matmul(x_gpu, x_gpu)
tensor_device(y_gpu)
cuda_provenance(y_gpu)

options(old)
```

`device = "cuda"` is strict: an unavailable or unhealthy runtime signals a
structured `cudaverse_cuda_unavailable` condition and never returns a CPU
result. `device = "auto"` may fall back, but provenance records the reason.

## Optional torch compatibility backend

Install R torch only when that compatibility path is wanted. Its LibTorch
runtime is separate from cudaverse and can be much larger than this package.
Follow the current official
[R torch installation guide](https://torch.mlverse.org/docs/articles/installation)
for supported platforms and CUDA builds.

```{r torch-order, eval = FALSE}
install.packages("torch")
options(cudaverse.cuda_backends = c("torch", "native"))
cuda_diagnostics()$backend_diagnostics$torch
```

The backend option constrains selection order for testing; it cannot bypass
runtime, capability, or self-test gates.

## Device and fallback semantics

| Request | Contract |
|---|---|
| `device = "cpu"` | Always use the portable CPU backend. |
| `device = "auto"` | Use the first eligible CUDA backend; otherwise use CPU with visible fallback provenance. |
| `device = "cuda"` | Require an eligible CUDA backend or signal a structured error; never silently use CPU. |

Use `tensor_device()` for tensors and `cuda_provenance()` for all current
results. Stage provenance distinguishes the requested policy, actual backend,
compute device, output device, selection reason, and host/device boundaries.
Graph and embedding workflows may contain intentional CPU or hybrid stages;
the [backend support article](backend-support.html) lists those boundaries.

## Memory and transfer guidance

Inspect memory without retaining a user tensor:

```{r memory-info}
memory <- cuda_memory_info("auto")
memory
```

The native report distinguishes whole-device `total_bytes`, `free_bytes`, and
`used_bytes` from cudaverse-owned `allocated_bytes` and
`allocated_peak_bytes`. The torch compatibility report exposes its allocator's
allocated and reserved counters. Unsupported counters stay `NA`; they are not
guessed from another backend's allocator. On the first CUDA selection in a
session, native peak bytes can include the released temporary allocations from
the runtime self-test.

- Exact `cuda_knn()` is quadratic in time. `batch_size` bounds the resident
  distance block; lower it when memory is constrained.
- `cuda_distance()` intentionally returns the complete dense pairwise matrix.
  Use `cuda_knn()` when only neighbours are required.
- `to_cpu()` and `to_dgCMatrix()` explicitly materialize host values.
- Native tensor subsetting and replacement are device operations for supported
  ordinary indices. Missing subscripts and compatibility backends may use a
  provenance-visible host path.
- PCA, distance blocks, stable top-k, native k-means, and supported sparse
  continuations reuse device storage. Inspect provenance rather than inferring
  residency from the requested device.
- Remove unused objects, call the relevant backend synchronization through the
  public operation that created them, then run `gc()` before diagnosing a
  suspected leak. Whole-device readings can include unrelated applications.

## Common failures

### Native is installed but not auto-eligible

Inspect `runtime_complete`, `missing_auto_capabilities`, and `self_test`. A
missing cuBLAS/cuSOLVER symbol, incompatible contract, or failed self-test is
fail-closed. Correct the runtime rather than forcing the backend past the gate.

### Explicit CUDA reports unavailable

Restart R after changing the NVIDIA driver or library paths, then rerun
`cuda_diagnostics()`. Confirm that `nvidia-smi` can see the device and that the
diagnostic paths identify the expected runtime files. Continue with
`device = "cpu"` until the strict smoke test succeeds.

### Results differ slightly across backends

Floating-point reduction and multiplication order can differ. Use the
operation's documented tolerance. kNN ties are resolved by original row order;
PCA comparisons should use reconstruction or subspace/projector error because
component signs are not unique.

### A result reports CPU or hybrid execution

This is not automatically a hidden fallback. Inspect every provenance stage.
UMAP, t-SNE, graph clustering, and parts of diffusion maps intentionally use
CPU packages in the current surface.

## Report a reproducible problem

Include the following without redacting backend failure reasons:

```{r report-template, eval = FALSE}
sessionInfo()
diagnostics <- cudaverse::cuda_diagnostics()
diagnostics$selected_backend
diagnostics$auto_selection_reason
diagnostics$backend_diagnostics
cudaverse::cuda_memory_info("auto")
cudaverse::cuda_provenance(result)
```

Also include a small input, the requested device, whether the CPU call passes,
the exact cudaverse commit or version, operating system, GPU model, driver
version, and the complete structured condition class/message. Do not begin
with a large production dataset.
